Geometric variational problems

### 2020-11-14 NSF2020/Description/background.tex
edited 4.1.2
fixed things here and there

The above considerations motivate the following definition for the HAP:
\begin{align}
\label{HAP}
  \Phi = \gamma \int_{\Omega} \rho |\nabla u|^2 + \rho^{-1}u^2 \,\dif V. 
\end{align}
The integrand contains
the decay length $\rho$
and a scalar function $u(\mathbf{x})$ called the water activity.
The integrand has units of an inverse length and so
the volume integral has units of area. Multiplication by interfacial tension
$\gamma$ makes $\Phi$ an energy.

@@

\begin{wrapfigure}[6]{l}{0.35\textwidth}
\centerline{\includegraphics[width=0.35\textwidth]{figures/Lysis.pdf}}
  \vspace{-8pt}
  \caption{\label{fig:lysis} \footnotesize Two hydrophobic particles
  enter and then lyse the circular bilayer.}
\end{wrapfigure}
The greatest strength of using HAP to model lipid bilayer membrane is
the ability to form discontinuities (interfacial singularity) from first
physical principles without any artificial manipulation to rearrange the
interface (Figures~\ref{fig:demixing} and~\ref{fig:lysis}). For
example, using the HAP model to simulate a vesicle under a shear flow
that is strong enough to cause the vesicle membrane to rupture (see
\S\ref{sec:preliminary_work}), we expect the HAP model to capture the
reorganization of lipid molecules on the scales of membrane thickness
($\sim 5$ nm), which is a nearly impossible task using phase-field or
immerse boundary method without extreme refinement around the membrane
and artificial treatment of reconnection during the topological change
of an interface \cite{doi:10.1063/5.0009734, LiAn-Chang16,
doi:10.1098/rspa.2012.0505, doi:10.1137/130941432, Feetzl18,
doi:10.1137/16M1108406}.

>>>>

The above considerations motivate the following definition for the HAP:
\begin{align}
\label{HAP}
  \Phi = \gamma \int_{\Omega} \rho |\nabla u|^2 + \rho^{-1}u^2 \,\dif V. 
\end{align}
The integrand contains
the decay length $\rho$
and a dimensionless scalar function $u(\mathbf{x})$ called the water activity.
The integrand has units of an inverse length and so
the volume integral has units of area. Multiplication by interfacial tension
$\gamma$ makes $\Phi$ an energy.

@@

\begin{wrapfigure}[6]{r}{0.35\textwidth}
\centerline{\includegraphics[width=0.35\textwidth]{figures/Lysis.pdf}}
  \vspace{-8pt}
  \caption{\label{fig:lysis} \footnotesize Two hydrophobic particles
  enter and then lyse the circular bilayer.}
\end{wrapfigure}
The greatest strength of using HAP to model lipid bilayer membrane is
the ability to form discontinuities (interfacial singularity) from first
physical principles without any artificial manipulation to rearrange the
interface (Figures~\ref{fig:demixing} and~\ref{fig:lysis}). For
example, using the HAP model to simulate a vesicle under a shear flow
that is strong enough to cause the vesicle membrane to rupture (see
\S\ref{sec:preliminary_work}), we expect the HAP model to capture the
reorganization of lipid molecules on the scales of membrane thickness
($\sim 5$ nm), which is a nearly impossible task using phase-field or
immersed boundary methods without extreme refinement around the membrane
and artificial treatment of reconnection during the topological change
of an interface \cite{doi:10.1063/5.0009734, LiAn-Chang16,
doi:10.1098/rspa.2012.0505, doi:10.1137/130941432, Feetzl18,
doi:10.1137/16M1108406}.

### 2020-11-14 NSF2020/Description/preliminarywork.tex
fixed things here and there

\begin{wraptable}[8]{r}{0.55\textwidth}
\centering{
  \begin{tabular}{|l|r|r|p{0.1\textwidth}|}
    \hline
          & HAP                     & experiment          & ref.\\
    \hline 
    $\KB$ & $11 \pm 2.5$ \kBT       & 10 \kBT               & \cite{Naetal15,VeBrPa15,NAGLE2000159,PhysRevLett.113.248102}\\
    \hline 
    $\KA$ & $34$ \kBT \; nm$^{-2}$  & 30--40 \kBT nm$^{-2}$ & \cite{Nagle17, Nagle17-2}\\
    \hline 
    $\KTH$ & $12$ \kBT \; nm$^{-2}$ & 10 \kBT nm$^{-2}$     & \cite{KUZMIN2005, KoNa15} \\ \hline
  \end{tabular}
}
  \vspace{-8pt}
\caption{\label{tab:moduli} \footnotesize Comparision of values elastic
  moduli from the experimental literature and values derived by HAP
  simulation.} 
\end{wraptable}
\section{Preliminary work}
\label{sec:preliminary_work}
The HAP project was initiated in 2017 by PIs RR and YNY. The lead author
on our first paper~\cite{Fu2018_SIAM} was Szu-Pei Fu, who was PI YNY's
PhD student and is currently a postdoc with PI RR. The collaboration
involved theoretical work and simulation. The theoretical work included
model development, proving the first variation formula~\eqref{stress}
and force-free law, and an energy-based uniqueness theorem for the
exterior domain problem~\eqref{SL}. Coauthors Andreas Kl\"ockner
(Computer Science, UIUC, NSF CAREER \#1654756) and his PhD student Matt
Wala provided their Quadrature by Expansion, or QBX algorithm and Fu ran
simulations on NJIT clusters.

@@

\begin{wrapfigure}[]{r}{0.5\textwidth}
  \centering{\includegraphics[width=0.5\textwidth]{Figures/TankTreading.pdf}}
  \vspace{-25pt}
    \caption{\label{fig:tank_treading} \footnotesize A two-dimensional
    particle-based vesicle in a shear flow. Panel A shows the particles,
    water activity, and background flow direction. Panel B plots the
    mean angular velocities for the outer leaflet and the inner
    respectively.  Panel C shows that the trajectory of the particle
    centers forms two ellipses with well-defined inclination angle
    $\Theta$. Panel D shows the evolution of the inclination angle.
    Panels B and D share the same time axis.}
\end{wrapfigure}
The PIs BQ, RR and YNY initiated a collaboration in Summer 2020 to
develop the hydrodynamic model of \S\ref{subsec:specific_aim_3}.  To
date, we have written solvers for the two-dimensional boundary integral
formulations of~\eqref{SL} and~\eqref{eq:stokes} that achieve
third-order accuracy up to the boundary. The specialized reciprocal
identity~\eqref{eq:reciprocal} gives spectral accuracy for force and
torque (Table~\ref{tab:spectral_force}).

We performed three types of simulations. The first measured bending by
loading a partially clamped planar bilayer. The second used a harmonic
bond to dilate a circular bilayer. The third measured tilt using a decay
equation from~\cite{KUZMIN2005}. These simulations isolated three of the
five deformations of~\eqref{ansatz3}, enabling us to read off elastic
moduli from simulation data. The results agreed remarkably well with the
values reported in the experimental literature (Table~\ref{tab:moduli}).
The agreement is underscored by the fact that the two main HAP
parameters, $\gamma$ and $\rho$, were assigned physical values from the
outset rather than being tuned to fit data.

We have simulated two-dimensional vesicles in shear flow
$\mathbf{u}_{\infty} = \dot{\gamma} x_2 \mathbf{i}_1$ with shear rate
$\dot{\gamma}$,
%(The symbol shear rate is unrelated to interfacial tension.)
and the particle-based system mimics the behavior observed in continuous
vesicles~\cite{torres-sanchez_millan_arroyo_2019,
mahapatra_saintillan_rangamani_2020, Steigmann99, C6SM02452A}. The basic
pattern is that the vesicle bilayer approaches a steady tank-treading
ellipse (Figure~\ref{fig:tank_treading}C). We analyzed the simulation
data in the context of theoretical work on
tank-treading~\cite{Finken2008}. Jeffrey's equation predicts a
tank-treading frequency $\dot{\gamma}/2$ (dashed line in
Figure~\ref{fig:tank_treading}B). In the particle simulation, the
tank-treading frequency corresponds to the mean of the particle angular
velocities, and these approach a value within 7\% of the theoretical
prediction (dotted and solid curves in Figure~\ref{fig:tank_treading}B).
In the low viscosity-contrast regime of our simulation, Jeffrey's
equation predicts a decrease in the ellipse's inclination angle $\Theta$
and our particle-based simulation follows this prediction
(Figure~\ref{fig:tank_treading}D). The higher frequency oscillation
comes from particle reorganization.

Analyzing for any underlying constitutive laws, the particles reorganize
so that there is an initial increase and then plateauing of vesicle
circumference. As is the case with real lipid bilayer membranes, the
particle collection permits a small amount of stretching. The
theoretical calculations give an increase of length by 3.6\% which
agrees with our measured increase in length. The simulations also
suggest that the particle system behaves as a semipermeable
vesicle~\cite{323e9a2f0c58487ea82518d7a1f96485, YAO2017728}.

@@

>>>>

\begin{wraptable}[8]{r}{0.55\textwidth}
\centering{
  \begin{tabular}{|l|r|r|p{0.1\textwidth}|}
    \hline
          & HAP                     & experiment          & ref.\\
    \hline 
    $\KB$ & $11 \pm 2.5$ \kBT       & 10 \kBT               & \cite{Naetal15,VeBrPa15,NAGLE2000159,PhysRevLett.113.248102}\\
    \hline 
    $\KA$ & $34$ \kBT \; nm$^{-2}$  & 30--40 \kBT nm$^{-2}$ & \cite{Nagle17, Nagle17-2}\\
    \hline 
    $\KTH$ & $12$ \kBT \; nm$^{-2}$ & 10 \kBT nm$^{-2}$     & \cite{KUZMIN2005, KoNa15} \\ \hline
  \end{tabular}
}
  \vspace{-8pt}
\caption{\label{tab:moduli} \footnotesize Comparision of values elastic
  moduli from the experimental literature and values derived by HAP
  simulation.} 
\end{wraptable}
\section{Preliminary work}
\label{sec:preliminary_work}
The HAP project was initiated in 2017 by PI RR and YNY. The lead author
on our first paper~\cite{Fu2018_SIAM} was Szu-Pei Fu, who was PI YNY's
PhD student and is currently a postdoc with PI RR. The collaboration
involved theoretical work and simulation. The theoretical work included
model development, proving the first variation formula~\eqref{stress}
and force-free law, and an energy-based uniqueness theorem for the
exterior domain problem~\eqref{SL}. Coauthors Andreas Kl\"ockner
(Computer Science, UIUC, NSF CAREER \#1654756) and his PhD student Matt
Wala provided their Quadrature by Expansion, or QBX algorithm and Fu ran
simulations on NJIT clusters.

@@

\begin{wrapfigure}[]{r}{0.5\textwidth}
  \centering{\includegraphics[width=0.5\textwidth]{Figures/TankTreading.pdf}}
  \vspace{-25pt}
  \caption{\label{fig:tank_treading} \footnotesize
    Panel A shows water activity and flow direction for  a two-dimensional
    particle-based vesicle in a shear flow; $\mu  = 1$ mPa s,
    $\dot{\gamma} = 0.5$ $\mu$s$^{-1}$. Panel B plots the
    mean angular velocities for the outer leaflet and the inner
    respectively.  Panel C shows that the trajectory of the particle
    centers forms two ellipses with well-defined inclination angle
    $\Theta$. Panel D shows the evolution of the inclination angle.
    Panels B and D share the same time axis.}
\end{wrapfigure}
The paper \cite{Fu2018_SIAM} considered three types of simulations. The first measured bending by
loading a partially clamped planar bilayer. The second used a harmonic
bond to dilate a circular bilayer. The third measured tilt using a decay
equation from~\cite{KUZMIN2005}. These simulations isolated three of the
five deformations of~\eqref{ansatz3}, enabling us to read off elastic
moduli from simulation data. The results agreed remarkably well with the
values reported in the experimental literature (Table~\ref{tab:moduli}).
The agreement is underscored by the fact that the two main HAP
parameters, $\gamma$ and $\rho$, were assigned physical values from the
outset rather than being tuned to fit data.

The PI BQ, RR and YNY initiated a collaboration in Summer 2020 to
develop the hydrodynamic model of \S\ref{subsec:specific_aim_3}.  To
date, we have written solvers for the two-dimensional boundary integral
formulations of~\eqref{SL} and~\eqref{eq:stokes} that achieve
third-order accuracy up to the boundary. The specialized reciprocal
identity~\eqref{eq:reciprocal} gives spectral accuracy for force and
torque (Table~\ref{tab:spectral_force}).

We have simulated two-dimensional vesicles in shear flow
$\mathbf{u}_{\infty} = \dot{\gamma} x_2 \mathbf{i}_1$ with shear rate
$\dot{\gamma}$,
%(The symbol shear rate is unrelated to interfacial tension.)
and the particle-based system mimics the behavior observed in continuous
vesicles~\cite{torres-sanchez_millan_arroyo_2019,
mahapatra_saintillan_rangamani_2020, Steigmann99, C6SM02452A}. The basic
pattern is that the vesicle bilayer approaches a steady tank-treading
ellipse (Figure~\ref{fig:tank_treading}C). We analyzed the simulation
data in the context of theoretical work on
tank-treading~\cite{Finken2008, PhysRevLett.106.158103}.
Jeffrey's equation predicts a
tank-treading frequency $-\dot{\gamma}/2$ (dashed line in
Figure~\ref{fig:tank_treading}B). In the particle simulation, the
tank-treading frequency corresponds to the mean of the particle angular
velocities, and these approach a value within 7\% of the theoretical
prediction (dotted and solid curves in Figure~\ref{fig:tank_treading}B).
In the low viscosity-contrast regime of our simulation, Jeffrey's
equation predicts a decrease in the ellipse's inclination angle $\Theta$
and our particle-based simulation follows this prediction
(Figure~\ref{fig:tank_treading}D), albeit with some higher frequency oscillation
coming from particle reorganization.

Analyzing for any underlying constitutive laws, the particles reorganize
so that there is an initial increase and then plateauing of vesicle
circumference. As is the case with real lipid bilayer membranes, the
particle collection permits a small amount of stretching. The
theoretical calculations give an increase of length by 3.6\% and this amount
agrees with our measured increase in length. The simulations also
suggest that the particle system behaves as a semipermeable
vesicle~\cite{323e9a2f0c58487ea82518d7a1f96485, YAO2017728}.

@@

In conclusion, the HAP model accurately mimics the behavior of continuous
membranes, while at the same time capturing the
reconnection during the topological change of 
lipid molecules on the scales of membrane thickness,
overcoming one of the long-standing challenges of continuum modeling.

### 2020-11-14 NSF2020/Description/specificaim1_ver2.tex
fixed things here and there

%  Past researchers expressed the strain tensor in terms of monolayer tangent plane.   
%  In this frame, however, there is a coupling between the gradients of the monolayer surface, lipid direction, and chain length,
%  and accounting for these couplings by Taylor expansion  
%  can lead to an energy density with an unruly number of parameters \cite{PhysRevE.102.042406}.
%
%The form of the elastic energy density \eqref{ansatz3} is the same as
%  the Oseen-Frank energy density for nematic liquid crystals \cite{ANDRIENKO2018520,Tran7106,Helfrich73}.   In fact,  
%  a lipid monolayer acts as one layer in a smectic  phase \cite{REYESMATEO1995978,Rangamani20140463,PhysRevLett.113.248102}. 
%  Motivated by liquid crystal theory, we have found that the gradient terms completely decouples when 
%  expanding the strain tensor, not in terms of the monolayer tangent plane, but in terms of a plane perpendicular to ${\bf n}$. 
%  Using this observation, we prove an exact identity 
%  that gives the incompressibility condition by a Steiner-type polynomial in $\Div {\bf n}$ and $\det \nabla {\bf n}$ \cite{Fe59}.
%  In contrast, the works \cite{TerziDeserno17, PhysRevE.102.042406, Hamm2000, C9SM02079A} utilize an
%  approximate identity for incompressibility, and so we can offer a considerably more accurate analysis of monolayer energy.
%  
%%To measure the twist modulus, we place a single monolayer on a hydrophobic substrate 
%%and take the initial configuration ${\bf n} = (0, f(x_1), \sqrt{1 - f^2(x_1)})$,
%%yielding a surface gradient $\nabla {\bf n} = \begin{pmatrix}0 & 0\\ f'(x_1) & 0\end{pmatrix}$. 
%%  Here $\Div {\bf n} = \mathrm{tr}(\nabla {\bf n}) = 0$, $\det(\nabla {\bf n}) = 0$ and
%%  $\Curl {\bf n} = (\nabla u)_{12} - (\nabla u)_{21} = f'(x_1)$.
%%  This gives the continuum elastic energy equals
%%  \begin{equation}
%%    \label{eq:twist_energy}
%%    E = \int_{\Sigma} \tfrac{1}{2}\KT (f'(x_1))^2 + \tfrac{1}{2} \KTH f^2(x_1) \,\dif dx_1 \dif x_2,
%%  \end{equation}
%%  which we compare to the HAP simulation data.
%
\subsubsection{Asymptotic analysis on the limit $\rho\rightarrow 0$ of HAP}
The HAP energy converges to the surface energy in the limit of vanishing
screening length. Also, the boundary layers of the solutions to the screen Laplace equation
converge to area measures supported on the particle boundaries.
This phenomena are analogous to the sharp interface limit in the phase-field functional
formulation~\cite{Modica87, MODICA1987487, LuMo89}. PI RR has worked
on phase field approximations of membrane bending
energy~\cite{0951-7715-18-3-016} and asymptotcs in the Poisson-Boltzman
equation~\cite{1531-3492_2006_2_357, Lee2018}.

@@

Another set of questions is to establish the so-called $\Gamma$-convergence of energy minimizing

>>>>

%  Past researchers expressed the strain tensor in terms of monolayer tangent plane.   
%  In this frame, however, there is a coupling between the gradients of the monolayer surface, lipid direction, and chain length,
%  and accounting for these couplings by Taylor expansion  
%  can lead to an energy density with an unruly number of parameters \cite{PhysRevE.102.042406}.
%
%The form of the elastic energy density \eqref{ansatz3} is the same as
%  the Oseen-Frank energy density for nematic liquid crystals \cite{ANDRIENKO2018520,Tran7106,Helfrich73}.   In fact,  
%  a lipid monolayer acts as one layer in a smectic  phase \cite{REYESMATEO1995978,Rangamani20140463,PhysRevLett.113.248102}. 
%  Motivated by liquid crystal theory, we have found that the gradient terms completely decouples when 
%  expanding the strain tensor, not in terms of the monolayer tangent plane, but in terms of a plane perpendicular to ${\bf n}$. 
%  Using this observation, we prove an exact identity 
%  that gives the incompressibility condition by a Steiner-type polynomial in $\Div {\bf n}$ and $\det \nabla {\bf n}$ \cite{Fe59}.
%  In contrast, the works \cite{TerziDeserno17, PhysRevE.102.042406, Hamm2000, C9SM02079A} utilize an
%  approximate identity for incompressibility, and so we can offer a considerably more accurate analysis of monolayer energy.
%  
%%To measure the twist modulus, we place a single monolayer on a hydrophobic substrate 
%%and take the initial configuration ${\bf n} = (0, f(x_1), \sqrt{1 - f^2(x_1)})$,
%%yielding a surface gradient $\nabla {\bf n} = \begin{pmatrix}0 & 0\\ f'(x_1) & 0\end{pmatrix}$. 
%%  Here $\Div {\bf n} = \mathrm{tr}(\nabla {\bf n}) = 0$, $\det(\nabla {\bf n}) = 0$ and
%%  $\Curl {\bf n} = (\nabla u)_{12} - (\nabla u)_{21} = f'(x_1)$.
%%  This gives the continuum elastic energy equals
%%  \begin{equation}
%%    \label{eq:twist_energy}
%%    E = \int_{\Sigma} \tfrac{1}{2}\KT (f'(x_1))^2 + \tfrac{1}{2} \KTH f^2(x_1) \,\dif dx_1 \dif x_2,
%%  \end{equation}
%%  which we compare to the HAP simulation data.
%
\subsubsection{Asymptotic analysis on the limit $\rho\rightarrow 0$ of HAP}
The HAP energy converges to the surface energy in the limit of vanishing
screening length~\cite{Shibata2004}.
Also, the boundary layers of the solutions to the screen Laplace equation
converge to area measures supported on the particle boundaries
\cite{Lee2018, Lin2015, Shibata2004}.
This phenomena are analogous to the sharp interface limit in the phase-field functional
formulation~\cite{Modica87, MODICA1987487, LuMo89}. PI RR has worked
on phase field approximations of membrane bending
energy~\cite{0951-7715-18-3-016} and asymptotcs in the Poisson-Boltzman
equation~\cite{1531-3492_2006_2_357, Lee2018}.

@@

Ultimately, we will be able to address the so-called $\Gamma$-convergence of energy minimizing
